Quantum program correctness is typically assured by formal verification, but the quantum semantic nature, based on unitary, density matrices, and complex numbers, indicates that such verification is laborious and time-comsuming. 
In this paper, we proposed quantum natural proof (QNP), an automated proof system for verifying classical quantum hybird algorithms. 
Natural proofs are a subclass of proofs that are amenable to completely automated reasoning, that provide sound but incomplete procedures, and that capture common reasoning tactics in program verification. 
The core of QNP is a quantum proof system, named the Qafny proof system, that maps quantum operations to classical array aggregate operations that can be effectively verified in a classical separation logic framework. We have shown the soundness and completeness of the Qafny proof system as well as the soundness of the proof system compilation from Qafny to Dafny.
Qafny permits quantum conditionals and we believe that our quantum conditional proof rule is the first quantifier free proof rule for quantum conditional operations.
By using the QNP implementation in Dafny, we enable the automated verification for many quantum algorithms, with both the saving of human efforts and computer execution time, including GHZ, Shor's algorithm, Grover's algorithm, and quantum walk algorithm.
In addition, quantum programs written in Qafny can be compiled to quantum circuits so that every verified quantum program can be run on a quantum machine. 
